---
title: "Milestone Report for Data Science Capstone Project"
author: "Jacob Spangler"
date: "Sunday, November 16, 2014"
output: html_document
---

# Summary
The Capstone project for the Coursera Data Science Specialization involves using the [HC Corpora][1] Dataset. The Capstone project is done in collaboration with [Swiftkey][2] and the goal of this project is to design a shiny application with text prediction capabilities. This report will outline the exploratory analysis of the dataset and the current plans for implementing the text prediction algorithm.

# Description of Data
The [HC Corpora][1] dataset is comprised of the output of crawls of news sites, blogs and twitter. A readme file with more specific details on how the data was generated can be found [here][3]. The dataset contains 3 files across four languages (Russian, Finnish, German and English). This project will focus on the English language datasets. The names of the data files are as follows:

1. en_US.blogs.txt
2. en_US.twitter.txt
3. en_US.news.txt

The datasets will be referred to as "Blogs", "Twitter" and "News" for the remainder of this report.

# Download the data

```{r, Downloading Data, eval=FALSE}
if(!file.exists("Coursera-SwiftKey.zip")){
    #Download the dataset
    download.file("https://d396qusza40orc.cloudfront.net/dsscapstone/dataset/Coursera-SwiftKey.zip",
                  "Coursera-SwiftKey.zip")
    Download_Date <- Sys.time()
    Download_Date
    #"2014-11-09 07:22:48 PST"
    
    #Unzip the dataset
    unzip("Coursera-SwiftKey.zip")
}else{
    print("Dataset is already downloaded!")
}
```

## Polish News Dataset
There is a minor problem with the News dataset. It contains an unusual character on line 77,259. In order to address this issue a small piece of code was written to edit out the character before processing the dataset.

```{r, Processing News, eval=FALSE}
if(!file.exists("./final/en_US/en_US.news_edit.txt")){
    con <- file("./final/en_US/en_US.news.txt", "rb")
    News_Data <- readLines(con)
    close(con)
    #Remove the odd symbol on line 77259
    News_Data <- gsub("\032", "", News_Data, ignore.case=F, perl=T)
    writeLines(News_Data, con=file("./final/en_US/en_US.news_edit.txt"))
    close(con=file("./final/en_US/en_US.news_edit.txt"))
    file.rename("./final/en_US/en_US.news.txt", "./final/en_US.news.txt")
}else{
    con <- file("./final/en_US/en_US.news_edit.txt", "rb")
    News_Data <- readLines(con) 
    close(con)
}
```

# Characteristics of Datasets

```{r, Read in Datasets, echo=TRUE, message=FALSE}
#Load libraries
library(NLP)
library(tm)
library(stringi)
library(ggplot2)
library(RWeka)
library(data.table)

#Generate Corpus for text analysis
cname <- file.path(".", "final", "en_US")
docs <- Corpus(DirSource(cname))
```

The first part of this exploratory analysis is to determine the basic characteristics for each dataset. These
characteristics are shown in the table below.

Dataset  | File Size (bytes) | Number of Lines | Smallest entry | Largest entry
------------- | ------------- | ------------- | ------------- | -------------
Blogs     | `r as.character(file.info("./final/en_US/en_US.blogs.txt")$size)`     | `r as.character(length(docs[[1]]$content))` | `r as.character(min(nchar(docs[[1]]$content)))` | `r as.character(max(nchar(docs[[1]]$content)))` 
Twitter   | `r as.character(file.info("./final/en_US/en_US.twitter.txt")$size)`  | `r as.character(length(docs[[3]]$content))` | `r as.character(min(nchar(docs[[3]]$content)))` | `r as.character(max(nchar(docs[[3]]$content)))` 
News      | `r as.character(file.info("./final/en_US/en_US.news_edit.txt")$size)`     | `r as.character(length(docs[[2]]$content))` | `r as.character(min(nchar(docs[[2]]$content)))` | `r as.character(max(nchar(docs[[2]]$content)))` 

## Subsetting and Processing the Dataset

Each of the datasets (Blogs, Twitter and News) are large enough that processing time is a factor. In order to address this concern, a representative sampling of each of the datasets was made for the remainder of this analysis. The subset of each file is outlined in the table below.

```{r, Subsetting and Polishing Datasets}
#Limit Dataset to a random subset of 20% of the data
set.seed(1337)
Subset <- docs
Subset[[1]]$content <- Subset[[1]]$content[as.logical(rbinom(length(Subset[[1]]$content),
                                                             1, prob=0.2))]
Subset[[2]]$content <- Subset[[2]]$content[as.logical(rbinom(length(Subset[[2]]$content),
                                                             1, prob=0.2))]
Subset[[3]]$content <- Subset[[3]]$content[as.logical(rbinom(length(Subset[[3]]$content),
                                                             1, prob=0.2))]
```

Dataset  | File Size (bytes) | Number of Lines | Smallest entry | Largest entry
------------- | ------------- | ------------- | ------------- | -------------
Subset Blogs     | `r as.character(object.size(Subset[[1]]$content))`     | `r as.character(length(Subset[[1]]$content))` | `r as.character(min(nchar(Subset[[1]]$content)))` | `r as.character(max(nchar(Subset[[1]]$content)))` 
Subset Twitter   | `r as.character(object.size(Subset[[3]]$content))`  | `r as.character(length(Subset[[3]]$content))` | `r as.character(min(nchar(Subset[[3]]$content)))` | `r as.character(max(nchar(Subset[[3]]$content)))` 
Subset News      | `r as.character(object.size(Subset[[2]]$content))`     | `r as.character(length(Subset[[2]]$content))` | `r as.character(min(nchar(Subset[[2]]$content)))` | `r as.character(max(nchar(Subset[[2]]$content)))` 

Before the subsetted data can be fully analyzed the data needs to be pre-processed to standardize the words and characters from each dataset. An example entry from the Blogs dataset is shown below:

```{r, Example Pre-Processing, echo=FALSE}
Subset[[1]]$content[10]
```

# Word Frequency
```{r, Document Term Matrix, echo=FALSE}
if(!file.exists("dtm.RData")){
    #Create a document term matrix
    dtm <- DocumentTermMatrix(Subset)
    save(dtm, file="dtm.RData")
}else{
    load("dtm.RData")
}

#Get word Frequencies
freq <- sort(colSums(as.matrix(dtm)), decreasing=TRUE)
wf <- data.frame(word=names(freq), freq=freq)
```

```{r, fig.height=4, fig.width=6}
#Plot Word Frequencies
ggplot(wf[wf$freq>60000, ], aes(x=word, y=freq)) +
    geom_bar(stat="identity") +
    theme(axis.text.x=element_text(angle=45, hjust=1)) +
    xlab("") +
    ylab("Frequency") +
    ggtitle("Words that appear over 60,000\ntimes in the three Datasets")
```

The high frequency for "connecting" words, such as "the", "and", "that" suggests that using a pattern based on word frequency alone will not be sufficient for text prediction. The next analysis looks at common word combinations.

# N-gram Frequency

For brivity the N-gram analysis of this report was limited to 2-grams.

```{r, Ngrams, echo=FALSE}
if(!file.exists("tdm.RData")){
    BigramTokenizer <- function(x) NGramTokenizer(x, Weka_control(min = 2, max = 4))
  	tdm <- TermDocumentMatrix(Subset, control = list(tokenize = BigramTokenizer))
  	save(tdm, file="tdm.RData")
}else{
	  load("tdm.RData")
}

NgramFreq <- sort(rowSums(as.matrix(tdm)), decreasing=TRUE)
WF_Ngram <- as.data.frame(data.table(word=names(NgramFreq), freq=NgramFreq))
```

```{r, fig.height=4, fig.width=6}
#Plot Word Frequencies
ggplot(WF_Ngram[WF_Ngram$freq>16000, ], aes(x=word, y=freq)) +
    geom_bar(stat="identity") +
    theme(axis.text.x=element_text(angle=45, hjust=1)) +
    xlab("") +
    ylab("Frequency") +
    ggtitle("Two Word Combinations (2-grams)that appear\n over 16,000 times in the three Datasets")
```

The distribution of 2-grams gives an idea of the prevalence of prepositions in natural language. The text prediction model will have to take this into account.

# Plan

The current plan for the development of the text prediction application will be to use the frequency of 4-grams, 3-grams and 2-grams to estimate the most likely word to follow the entered text. The trick will be to offer valid predictions of N-grams that are not observed within the dataset. In these cases the algorithm will likely default to a list of "non-common" words (i.e. factor out words like the, and, that) and estimate the best possible candidate.

# Session Information
This analysis was performed on a machine with the following characteristics:

```{r, Session Info, echo=FALSE}
sessionInfo()
```

[1]: http://www.corpora.heliohost.org/ "HC Corpora"
[2]: http://swiftkey.com/en/ "Swiftkey"
[3]: http://www.corpora.heliohost.org/aboutcorpus.html "Corpora Readme"
